---
title: "Identify and Sample Multimodal Genes"
author: "Jeremy Sieker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Identify and Sample Multimodal Genes"}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
fig_width: 10 
fig_height: 4 
---
___
<center> <h1>Introduction</h1> </center>
  The main goal of this package is to aid in quickly analyzing thousands of genes' distributions. By doing this, we can pick out groups of genes whose expressions follow patterns of interest identifiable through their multimodality. In order to illustrate this process, we will import and prepare a RNAseq dataset, identify some logical places to break up the distributions, and take a closer look at some of the genes we identify in that process.  
  
  \vspace{5mm}
  To start, we will import a data set from Expression Atlas' public resources.
   \vspace{5mm}
```{r}
library(DipEx)
ss <- read.table(url("https://www.ebi.ac.uk/gxa/experiments-content/E-GEOD-70484/resources/BaselineProfilesWriterService.RnaSeq/tsv"), sep = '\t', header = TRUE)
ss[1:5,1:3]
```
   \vspace{5mm}
  Ideally, we would have the gene names as the row names and there would be nothing except gene data in the actual frame, so we will do a few clean-ups and also cut the number of samples in half for quicker computation.
   \vspace{5mm}
```{r}
ss <- ss[(as.numeric(row.names(ss)) %% 2 == 1),]
row.names(ss) <- ss$Gene.ID
ss <- ss[,-c(1:2)] #removing everything that is not expression data
ss[1:5,1:2]
```
 \vspace{5mm}
We will begin our analysis by getting an idea of how the dip values in this set are distributed. By doing this, we can get an idea of where we should set our region break points.
 \vspace{5mm}
```{r, fig.width=7.25, fig.align='center'}
mod1 <- previewDipDistribution(RNAdata = ss, barLine = 0.025)
plot(mod1$DipPlot)
```
 \vspace{5mm}
Based on this plot, it seems that it could be reasonable to place break points at Dip = 0.05 and Dip = 0.11. Given the required syntax of the break input, the way to define this is as follows:
```{r}
b <- c(-Inf, 0.05, 0.11, Inf)
```
 \vspace{5mm}
Having defined our break regions, our next step is to get an idea of what the gene distributions in each of these regions look like. To plot a couple random selections from each region, we could use the following code:
 \vspace{5mm}
```{r, fig.width=5.5, fig.height=4, fig.align='center'}
mod2 <- plotSamplesByDipRegion(RNAdata = ss, breaks = b, samples = 4, xlab = "Sample RNA Expression")
```
 \vspace{5mm}
At this point, 'mod2' is a list. To access any of these plots, you can use the code 'plot(mod2$PlotX)' with the X replaced with the region number whose samples you are trying to plot. 

\vspace{5mm}
In order to give an example of how to analyze down to the single gene level, let's focus in on one region. If you look up at the Region 1 plots we just made, a couple (or all) of the Region 1 samples look like perfect normal distributions around zero. These genes are the ones that had zero counts in all samples (in DESeq2-normalized samples, they will be around -0.29, instead). However, that doesn't mean that Region 1 is all the same, because there are typically some genes in the lowest dip region that had counts in one or a few samples. To access those, one could subset the DF output of mod2 to just get values from Region 1. Then to identify those that have counts, subset in such a way as to identify genes with a Zero-Excluded Median greater than zero.
 \vspace{5mm}
```{r, results = "hide"}
table(mod2$DF$Region)
mod2.r1 <- mod2$DF[(mod2$DF$Region==1),]
mod2.r1.g0 <- mod2.r1[(mod2.r1$ZeroXMedian > 0),]
```
\vspace{5mm}
If these were the genes you had in mind, you could then visualize a random sample of them as follows:
\vspace{5mm}
```{r, fig.width = 5.5, fig.height = 4, fig.align = 'center'}
sam <- mod2.r1.g0$GeneID[sample(nrow(mod2.r1.g0), size = 4, replace = FALSE)]
mod3 <- plotSamplesByName(nameList = sam, RNAdata = ss)
```
\vspace{5mm}
Alternatively, you could extract their Gene ID's with simple subsetting and then export them for whichever kind of further analysis you prefer. Any genes (up to 8 at a time) can be visualized in this manner using plotSamplesByName. To give another example, let's look at some genes on the other end of the dip spectrum. Not just those in Region 3, but those at the far end of region 3 (Dip values exceeding 0.20)

```{r, fig.width = 5.5, fig.height = 4, fig.align = 'center'}
mod2.high <- mod2$DF[(mod2$DF$DipOutput>=0.20),]
sam2 <- mod2.high$GeneID[sample(nrow(mod2.high), size = 4, replace = FALSE)]
mod3.1 <- plotSamplesByName(nameList = sam2, RNAdata = ss)
```
Looking at these plots, it appears that these genes are expressed very differently in different tissues, as compared to genes that are more normally-distributed. These different patterns of expression could have meaning as to the corresponding genes' functions/roles, so the simplicity of doing targeted searches like these could make a great tool.

If you are interested more in the region separation than the plots, there are a couple of ways to access that sort of data. The most direct method of getting each gene's Dip value and Region is with the following function:
 \vspace{5mm}
```{r, message=FALSE}
mod3 <- dipExtension(RNAdata = ss, breaks = b)
head(mod3)
```
Alternatively, a dataframe of nearly identical content and structure is accessible through the plotSamplesByDipRegion function we used earlier to make our first sample plots:
\vspace{5mm}
```{r}
head(mod2$DF)
```



